\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-30 coqimpl.tex
Reply comments. Compile is very different from Seal.
Code Highlighting for Family Use lstlisting insteadm but I cannot control the line spacing.

\textbf{Family compiled into module.} We support family by compiling
family components into Coq modules, family ``types'' into module types, and the context of a given term
into parameters of the module. We need to faithfully
reflect the \textit{late binding} nature of families: for each defining
field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using
``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall
\cL \sigma. T$. However, instead of using Coq's universal quantifier, we
choose module and module parameters to achieve this wrapping---each
family type in the context of the judgement will be one module parameter
with the compiled module type, and thus each field of the family will
compile into a functor parametrized by its context. Doing so we can also
get rapid feedback from type checking when the users are defining each
field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b : a = a
  := eq_refl.
Final Field a' : nat 
  := 1. 
EndFamily.
\end{minted}
  \end{minipage}%\YZ{Can the self's be omitted?}\EDJreply{Yes. But in our current plugin implementation it is not omitted. I think altering example is fine. If you think some small distance between example and the current implementation is fine, please comment back and I will remove all the self_}\YZreply{If omitting the self's poses no technical challenges, then I'd say it's fine to also omit them in this figure.}\EDJreply{Omitting self_ is easy and not much engineering effort is required. But omitting self_ will lead to `A.a' instead of `a' actually. So if we really want `a' as you last time mentioned, then a resolution is required and more engineering effort is required. But I don't see any technical difficulties. I have removed all the self, please check.}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

\begin{figure}[!htb]
  \begin{minipage}[t]{0.45\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
Family I.
 Field a : nat := 1.
 Field b : a = a := eq_refl.
 Fail Field c : a = 1 := eq_refl. (*Fail 1*)

@@

 Family MonotonicF.
  Field f : nat -> nat := non_impl.
  Field p : monotonic f := non_impl.
 SealFamily. 
EndFamily.
\end{minted}
  \end{minipage}
\begin{minipage}[t]{0.45\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
Family I2 extends I.
 Override Field a : nat := 1. (* Override *)
 Inherit b.                   (* Inherit  *)

@@

 Inherits Until MonotonicF.
 Override Family MonotonicF.
  Field f : nat -> nat := (+ 1).
  Field p : monotonic f := ...
 EndFamily. 
EndFamily.
\end{minted}
  \end{minipage}
  \caption{Example Code for Inheritance and Reasoning}\label{fig:plugin-example3}
\end{figure}

@@

\begin{figure}[!htb]
  \begin{minipage}[t]{0.32\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
Family B.
  FInductive b : Set 
    := tt : b | ff : b.
  Family neg_handler. 
    Field tt : self_B.b 
      := self_B.ff.
    Field ff : self_B.b 
      := self_B.tt.
  EndFamily.
  FRecursor neg 
    about self_B.b 
    motive (fun _ => self_B.b)
    using self_B.neg_handler
    by _rec.
  Field example := 
    self_B.neg self_B.tt. 
EndFamily.

Family B2 extends B.
  Extend FInductive b : Set 
    := uu : b.
  Extend Family neg_handler.
    Field uu : self_B2.b 
      := self_B2.uu.
  EndFamily. 
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.65\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
  (* Abstraction for Inductive Type *)
Module Type b_3 (self_B: EmptySig_4).
Parameter (b : Set). Parameter (tt ff : b). End b_3.      
Module Type b_6 := b_3 EmptyMod.
  (* Recursor to Set for b *)
Module b_rec_12 (self_B: b_6).
Definition __recursor_type_b_rec :=
  forall P : self_B.b -> Set,
  P self_B.tt -> P self_B.ff 
  -> forall __i : self_B.b, P __i. End b_rec_12. 
  (* Field B.neg_handler.ff *)
Module ff_24 (self_B: b_6) 
             (self_neg_handler: tt_23 self_B).
Definition ff : self_B.b := self_B.tt. End ff_24.
  (* Intermediate Module solely for type checking *)
Module v_33_34 (self_B: neg_handler_26).
Include b_rec_12 self_B.
Parameter (recursor_for_type_checking : __recursor_type_b_rec).
Definition term_for_type_checking :=
  recursor_for_type_checking (fun _ : self_B.b => self_B.b)
    self_B.neg_handler.tt
    self_B.neg_handler.ff.  
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Example Code for Inductive Type}\label{fig:plugin-example2}
\end{figure}

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
Unfortunately, there are cases an extensible proof brings more boilerplate code. For example, in STLC, we expect
``\mintinline{Coq}{~ value (tm_app x y)}'' to hold in all future
extensions of STLC---applications will never be considered as values
by any extension.
We know the correct way to prove the proposition
\mintinline{Coq}{value_not_app : value (tm_app x y) -> False.} is to use
"FRecursor" to inductively reason "value (tm_app x y)". However, this
means that every time we extend "value" with new value forms, we will
need to extend corresponding inductive cases for "app_not_value". We can
imagine the newly extended proof to be boring case analysis that the
newly added value forms are not "tm_app". What's worse, similar
scenario can happen on other statement like \mintinline{Coq}{~ value
(tm_if cond x y)}. Everything here is extended correctly but we end up
having boilerplate code, \textbf{semantically}.
\begin{figure}[!htb]
\begin{minted}{Coq}
Family STLC.
  FInductive tm : Set := ... 
  FInductive value : tm -> Prop := ... 
  Closing Fact value_not_app : forall x y, ~ value (tm_app x y) 
      by { intros x y H; inversion H; eauto }.
  Fail Ltac inv := 
    match goal with 
    | [h : value (tm_app _ _) |- ] => destruct (value_not_app _ _ h)
    (* Fail: value, tm_app, value_not_app unfound *)
    end. 
  MetaData tactic1.
    Ltac inv := (* ... the same definition as above ... *)
  EndMetaData.
EndFamily.
\end{minted}  
\caption{Example for Global Reasoning and MetaData}\label{fig:plugin-example-global-reasoning-meta-data}
\end{figure}

@@

\ref{prop:detectable-partiality} is derivable for all the finitary branching inductive type\footnote{By simply enumerating through the finite domain "Bᵢ a"}---that means all of the examples shown in this paper. \ref{prop:detectable-partiality} is not supported when using function (infinitary branching) in the constructors, for example, the case of using HOAS to represent functions in "tm". In those cases, we still have the fact that partial recursor is a conservative extension, but we cannot show that partial recursor is the extensional characterization any more。

>>>>

\textbf{Family compiled into module.} We support family by compiling
family components into Coq modules, family ``types'' into module types, and the context of a given term
into parameters of the module. We need to faithfully
reflect the \textit{late binding} nature of families: for each defining
field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using
``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall
\cL \sigma. T$. However, instead of using Coq's universal quantifier, we
choose module and module parameters to achieve this wrapping---each
family type in the context of the judgement will be one module parameter
with the compiled module type, and thus each field of the family will
compile into a functor parametrized by its context. Doing so we can also
get rapid feedback from type checking when the users are defining each
field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.30\linewidth}
\begin{lstlisting}[language=Coq, escapeinside={@}{@}]
Family A.
Field a : nat := 1.
Field b : a = a := eq_refl.
Final Field a' : nat := 1. 
EndFamily.
\end{lstlisting}
  \end{minipage}%\YZ{Can the self's be omitted?}\EDJreply{Yes. But in our current plugin implementation it is not omitted. I think altering example is fine. If you think some small distance between example and the current implementation is fine, please comment back and I will remove all the self_}\YZreply{If omitting the self's poses no technical challenges, then I'd say it's fine to also omit them in this figure.}\EDJreply{Omitting self_ is easy and not much engineering effort is required. But omitting self_ will lead to `A.a' instead of `a' actually. So if we really want `a' as you last time mentioned, then a resolution is required and more engineering effort is required. But I don't see any technical difficulties. I have removed all the self, please check.}
  \begin{minipage}[t]{0.35\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

\begin{figure}[!htb]
  \begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[language=Coq, escapeinside={@}{@}]
Family I.
 Field a : nat := 1.
 Field b : a = a := eq_refl.
 Fail Field c : a = 1 := eq_refl. (*Fail 1*)

@@

 Family MonotonicF.
  Field f : nat -> nat := non_impl.
  Field p : monotonic f := non_impl.
 SealFamily. 
EndFamily.
\end{lstlisting}
  \end{minipage}
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family I2 extends I.
 Override Field a : nat := 1. (* Override *)
 Inherit b.                   (* Inherit  *)

@@

 Inherits Until MonotonicF.
 Override Family MonotonicF.
  Field f : nat -> nat := (+ 1).
  Field p : monotonic f := ...
 EndFamily. 
EndFamily.
\end{lstlisting}
  \end{minipage}
  \caption{Example Code for Inheritance and Reasoning}\label{fig:plugin-example3}
\end{figure}

@@

\begin{figure}[!htb]
  \begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family B.
  FInductive b : Set 
    := tt : b | ff : b.
  Family neg_handler. 
    Field tt : self_B.b 
      := self_B.ff.
    Field ff : self_B.b 
      := self_B.tt.
  EndFamily.
  FRecursor neg 
    about self_B.b 
    motive (fun _ => self_B.b)
    using self_B.neg_handler
    by _rec.
  Field example := 
    self_B.neg self_B.tt. 
EndFamily.
Family B2 extends B.
  Extend FInductive b : Set 
    := uu : b.
  Extend Family neg_handler.
    Field uu : self_B2.b 
      := self_B2.uu.
  EndFamily. 
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.65\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
  (* Abstraction for Inductive Type *)
Module Type b_3 (self_B: EmptySig_4).
Parameter (b : Set). Parameter (tt ff : b). End b_3.      
Module Type b_6 := b_3 EmptyMod.
  (* Recursor to Set for b *)
Module b_rec_12 (self_B: b_6).
Definition __recursor_type_b_rec :=
  forall P : self_B.b -> Set,
  P self_B.tt -> P self_B.ff 
  -> forall __i : self_B.b, P __i. End b_rec_12. 
  (* Field B.neg_handler.ff *)
Module ff_24 (self_B: b_6) 
             (self_neg_handler: tt_23 self_B).
Definition ff : self_B.b := self_B.tt. End ff_24.
  (* Intermediate Module solely for type checking *)
Module v_33_34 (self_B: neg_handler_26).
Include b_rec_12 self_B.
Parameter (recursor_for_type_checking : __recursor_type_b_rec).
Definition term_for_type_checking :=
  recursor_for_type_checking (fun _ : self_B.b => self_B.b)
    self_B.neg_handler.tt
    self_B.neg_handler.ff.  
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Example Code for Inductive Type}\label{fig:plugin-example2}
\end{figure}

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
Unfortunately, there are cases an extensible proof brings more boilerplate code. For example, in STLC, we expect
``\mintinline{Coq}{~ value (tm_app x y)}'' to hold in all future
extensions of STLC---applications will never be considered as values
by any extension.
We know the correct way to prove the proposition
\mintinline{Coq}{value_not_app : value (tm_app x y) -> False.} is to use
"FRecursor" to inductively reason "value (tm_app x y)". However, this
means that every time we extend "value" with new value forms, we will
need to extend corresponding inductive cases for "app_not_value". We can
imagine the newly extended proof to be boring case analysis that the
newly added value forms are not "tm_app". What's worse, similar
scenario can happen on other statement like \mintinline{Coq}{~ value
(tm_if cond x y)}. Everything here is extended correctly but we end up
having boilerplate code, \textbf{semantically}.
\begin{figure}[!htb]
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
  FInductive tm : Set := ... 
  FInductive value : tm -> Prop := ... 
  Closing Fact value_not_app : forall x y, ~ value (tm_app x y) 
      by { intros x y H; inversion H; eauto }.
  Fail Ltac inv := 
    match goal with 
    | [h : value (tm_app _ _) |- ] => destruct (value_not_app _ _ h)
    (* Fail: value, tm_app, value_not_app unfound *)
    end. 
  MetaData tactic1.
    Ltac inv := (* ... the same definition as above ... *)
  EndMetaData.
EndFamily.
\end{lstlisting}  
\caption{Example for Global Reasoning and MetaData}\label{fig:plugin-example-global-reasoning-meta-data}
\end{figure}

@@

\ref{prop:detectable-partiality} is derivable for all the finitary branching inductive type\footnote{By simply enumerating through the finite domain "Bᵢ a"}---that includes all of the examples shown in this paper. \ref{prop:detectable-partiality} is not supported when using function (infinitary branching) in the constructors, for example, the case of using \textit{higher order abstract syntax} to represent functions in "tm". In those cases, we still have the fact that partial recursor is a conservative extension, but we cannot show that partial recursor is the extensional characterization any more。

### 2022-06-30 lang-design.tex
Code Highlighting for Family Use lstlisting insteadm but I cannot control the line spacing.

\begin{figure}[!htb]
  \begin{minipage}[t]{0.47\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=||]{Coq}
Family STLC.|\YZ{Does minted support adding extra keywords such as 'Family' and 'extends'?}|
Inductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
Inductive term : Set := 
  | var : id -> term 
  | lam : id -> term -> term ...
  | tunit : term 
Fixpoint subst 
  : id -> term -> term := ...
Inductive has_type 
  : context -> term -> type := ...
Proposition subst_lemma :
  has_type (Γ; x ↦ A) t T ->
  has_type Γ u A ->
  has_type Γ (subst x u t) T.
Inductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family STLC_bool extends STLC.
Extend Inductive ty : Set :=
  | bool : ty.

@@

Extend Inductive step : term -> term -> Prop 
  := ... (* Need to expand this binary relation *)
(* ... and more extension *)
EndFamily.
\end{minted}
  \end{minipage}
  \caption{Example STLC and its extension}\label{fig:STLC-example}
\end{figure}

>>>>

\begin{figure}[!htb]
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
Inductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
Inductive term : Set := 
  | var : id -> term 
  | lam : id -> term -> term 
  | tunit : term ...
Fixpoint subst 
  : id -> term -> term := ...
Inductive has_type 
  : context -> term -> type := ...
Proposition subst_lemma :
  has_type (G; x ↦ A) t T ->
  has_type G u A ->
  has_type G (subst x u t) T.
Inductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC_bool extends STLC.
Extend Inductive ty : Set :=
  | bool : ty.

@@

Extend Inductive step : term -> term -> Prop 
  := ... (* Need to expand this binary relation *)
(* ... and more extension *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \caption{Example STLC and its extension}\label{fig:STLC-example}
\end{figure}
